Compositional Certified Resource Bounds

### 2015-04-13 paper/main.tex
more rule descriptions

The rules {\sc Q:IncP}\iffull{, {\sc Q:DecP},}{} and {\sc Q:Inc}
describe how the potential is distributed after a size change of
a variable.
\iffull{The rule {\sc Q:IncP} is for increments $x \gets x + y$
and {\sc Q:DecP} is for decrements $x \gets x - y$, they both apply only
when we can deduce from the logical context $\Gamma$ that $y \ge 0$.
Of course, we have symmetrical rules {\sc Q:IncN} and {\sc Q:DecN}
that can be applied if $y$ is not positive.}{The rule {\sc Q:IncP} is for increments
$x \gets x+y$, it applies when we can deduce from the logical context
$\Gamma$ that $y \ge 0$.  Of course, we have symmetrical rules that
can be applied for decrements and when $y$ is not positive.}
The rules are all equivalent in the case where $y=0$.
The rule {\sc Q:Inc} can be applied
if we cannot deduce any information about $y$.  In $\toolname$,
the rules for increments and decrements are combined into one rule\iffull{ where the specialized rules take priority over {\sc Q:Inc}.}{.}

\iffull{To explain how rules for increment and decrement work, it is
sufficient to understand the rule {\sc Q:IncP}.  The others follow the
same idea and are symmetrical.}{}  In {\sc Q:IncP}, the program updates a
variable $x$ with $x+y$ where $y \ge 0$.  Since $x$ is changed, the
quantitative annotation must be updated to reflect the change of the
program state.  We write $x'$ for the value of $x$ after the assignment.
Since $x$ is the only variable changed, only intervals of the form
$\inter u x$ and $\inter x u$ will be resized. Note that for any $u$,
$\inter x u$ will get smaller with the update, and if $x' \in \inter x
u$ we have $|\inter x u| = |\inter x {x'}| + |\inter {x'} u|$.  But
$|\inter x {x'}| = |\inter 0 y|$ which means that the potential
$q'_{0y}$ in the postcondition can be increased by $q_{xu}$ under the
guard that $x' \in \inter x u$.  Dually, the interval $\inter v x$ can
get bigger with the update.  We know that $|\inter v {x'}| \le y +
|\inter v x|$.  So we decrease the potential of $\inter 0 y$ by
$q_{vx}$ to pay for this change.  The rule ensures this
only for $v \not\in \mathcal U$ because $x \le v$
otherwise, and thus $|\inter v x| = 0$.

\iffull{
Another interesting rule is {\sc Q:Call}.  It needs to account for the
changes to the stack caused by the function call, the arguments/return
value passing, and the preservation of local variables.  We can sum up
the main ideas of the rule as follows.
\itemskip
\begin{itemize}
\item The potential in the pre- and postcondition of the function
  specification is equalized to its matching potential in the callee's
  pre- and postcondition.
\itemskipIn
\item
  $\!$The potential of intervals $|\inter x y|$ is preserved
  if $x$ and $y$ are local\iffull{ variables.}{.}
\item
  The unknown potentials after the call (e.g.
  $|\inter x g|$, with $x$ local and $g$ global)
  are set to zero in the postcondition.
\end{itemize}
\itemskip
If $x$ and $y$ are local variables and $f(x,y)$ is
called, {\sc Q:Call} splits the potential of
$|\inter x y|$ in two parts, one part to perform the
computation in the function $f$ and one part to keep for
later use after the function call.  This splitting is
realized by the equations $Q = P {+} S$ and $Q' =
P' {+} S'$.  Arguments in the function precondition
$(\Gamma_f, Q_f)$ are named using a fixed vector $\Vargs$
of names different from all program variables.  This
prevents name conflicts from happening and ensures that
the substitution $[\Vargs/\vec x]$ is meaningful.
Symmetrically, we use the unique name $\Vret$ to represent
the return value in the function's postcondition
$(\Gamma'_f, Q'_f)$.
}{}

The rule {\sc Q:Loop} is a cornerstone of our analysis.  To apply it on a
loop body one needs to find an invariant potential $Q$ that will pay for
the iterations.  At each iteration, $M_l$ resources are spent to jump back,
this explains the postcondition $Q{+}M_l$. %  Assuming $M_l>0$, and if $Q$
%is finite on the initial heap (i.e. the precondition is not $\bot$), the
%loop can only run for a finite number of iterations.
To exit the loop,
the programmer uses the \code{break} statement, this explains why the postcondition
for the full $\code{loop}~S$ statement $\{ \Gamma'; Q' \}$ is used as break
postcondition in the derivation for $S$.

\iffull{The rule {\sc Q:Weak} is not syntax directed. In the implementation
we apply {\sc Q:Weak} before loops and between the two statements of a
sequential composition. We could integrate weakening into every
syntax directed rule but this simple heuristic helps to make the analysis
efficient.}{}  The high-level idea of {\sc Q:Weak} is the following: If
we have a sound judgement, then it is sound to add more potential to the
precondition and remove potential from the postcondition.  The concept of
\emph{more potential} is formalized by the relation $Q' \succeq_\Gamma Q$
that is defined in the rule {\sc Relax}\iffull{ that
also deals with the important task of transferring constant potential
(represented by $q_0$) to interval sizes and vice versa.  If we can deduce
from the logical context that the interval size $|\inter x y| \geq l$ is
larger then a constant $l$ then we can transfer potential $q_{xy} {\cdot}
|\inter x y|$ form the interval to constant potential $l{\cdot}q_{xy}$
and guarantee that we do not gain potential.  Conversely, if $|\inter
x y| \leq u$ for a constant $u$ then we can transfer constant potential
$u{\cdot}q_{xy}$ to the interval potential $q_{xy} {\cdot} |\inter x y|$ without gaining potential.}{.}

>>>>

The rules {\sc Q:IncP}, {\sc Q:DecP}, and {\sc Q:Inc}
describe how the potential is distributed after a size change of
a variable.
\iffull{The rule {\sc Q:IncP} is for increments $x \gets x + y$
and {\sc Q:DecP} is for decrements $x \gets x - y$, they both apply only
when we can deduce from the logical context $\Gamma$ that $y \ge 0$.
Of course, we have symmetrical rules {\sc Q:IncN} and {\sc Q:DecN}
that can be applied if $y$ is not positive.}{The rule {\sc Q:IncP} is for increments
$x \gets x+y$, it applies when we can deduce from the logical context
$\Gamma$ that $y \ge 0$.  Of course, we have symmetrical rules that
can be applied for decrements and when $y$ is not positive.}
The rules are all equivalent in the case where $y=0$.
The rule {\sc Q:Inc} can be applied
if we cannot find the sign of $y$.  % In $\toolname$,
% the rules for increments and decrements are combined into one rule\iffull{ where
% the specialized rules take priority over {\sc Q:Inc}.}{.}

To explain how rules for increment and decrement work, it is
sufficient to understand the rule {\sc Q:IncP}.  The others follow the
same idea and are symmetrical. In {\sc Q:IncP}, the program updates a
variable $x$ with $x+y$ where $y \ge 0$.  Since $x$ is changed, the
quantitative annotation must be updated to reflect the change of the
program state.  We write $x'$ for the value of $x$ after the assignment.
Since $x$ is the only variable changed, only intervals of the form
$\inter u x$ and $\inter x u$ will be resized. Note that for any $u$,
$\inter x u$ will get smaller with the update, and if $x' \in \inter x
u$ we have $|\inter x u| = |\inter x {x'}| + |\inter {x'} u|$.  But
$|\inter x {x'}| = |\inter 0 y|$ which means that the potential
$q'_{0y}$ in the postcondition can be increased by $q_{xu}$ under the
guard that $x' \in \inter x u$.  Dually, the interval $\inter v x$ can
get bigger with the update.  We know that $|\inter v {x'}| \le y +
|\inter v x|$.  So we decrease the potential of $\inter 0 y$ by
$q_{vx}$ to pay for this change.  The rule ensures this
only for $v \not\in \mathcal U$ because $x \le v$
otherwise, and thus $|\inter v x| = 0$.

The rule {\sc Q:Loop} is a cornerstone of our analysis.  To apply it on a
loop body one needs to find an invariant potential $Q$ that will pay for
the iterations.  At each iteration, $M_l$ resources are spent to jump
back, this explains the postcondition $Q{+}M_l$.  To exit the loop,
the programmer uses the \code{break} statement, it is the reason why
the postcondition for the full $\code{loop}~S$ statement $\{ \Gamma';
Q' \}$ is used as break postcondition in the derivation for $S$.

Another interesting rule is {\sc Q:Call}.  It needs to account for the
changes to the stack caused by the function call, the arguments/return
value passing, and the preservation of local variables.  We can sum up
the main ideas of the rule as follows.
\itemskip
\begin{itemize}
\item The potential in the pre- and postcondition of the function
  specification is equalized to its matching potential in the callee's
  pre- and postcondition.
\itemskipIn
\item
  The potential of intervals $|\inter x y|$ is preserved
  across a function call
  if $x$ and $y$ are local\iffull{ variables.}{.}
\item
  The unknown potentials after the call (e.g.
  $|\inter x g|$, with $x$ local and $g$ global)
  are set to zero in the postcondition.
\end{itemize}
\itemskip
If $x$ and $y$ are local variables and $f(x,y)$ is
called, {\sc Q:Call} splits the potential of
$|\inter x y|$ in two parts, one part to perform the
computation in the function $f$ and one part to keep for
later use after the function call.  This splitting is
realized by the equations $Q = P {+} S$ and $Q' =
P' {+} S'$.  Arguments in the function precondition
$(\Gamma_f; Q_f)$ are named using a fixed vector $\Vargs$
of names different from all program variables.  This
prevents name conflicts from happening and ensures that
the substitution $[\Vargs/\vec x]$ is meaningful.
Symmetrically, we use the unique name $\Vret$ to represent
the return value in the function's postcondition
$(\Gamma'_f; Q'_f)$.

The rule {\sc Q:Weak} is not syntax directed. In the implementation
we apply {\sc Q:Weak} before loops and between the two statements of a
sequential composition. We could integrate weakening into every
syntax directed rule but this simple heuristic helps to make the analysis
efficient.  The high-level idea of {\sc Q:Weak} is the following: If
we have a sound judgement, then it is sound to add more potential to the
precondition and remove potential from the postcondition.  The concept of
\emph{more potential} is formalized by the relation $Q' \succeq_\Gamma Q$
that is defined in the rule {\sc Relax} that
also deals with the important task of transferring constant potential
(represented by $q_0$) to interval sizes and vice versa.  If we can deduce
from the logical context that the interval size $|\inter x y| \geq l$ is
larger than a constant $l$ then we can turn the potential $q_{xy} {\cdot}
|\inter x y|$ form the interval into the constant potential $l{\cdot}q_{xy}$
and guarantee that we do not gain potential.  Conversely, if $|\inter
x y| \leq u$ for a constant $u$ then we can transfer constant potential
$u{\cdot}q_{xy}$ to the interval potential $q_{xy} {\cdot} |\inter x y|$
without gaining potential.
